---
format: 
  html: 
    theme: cyborg
    toc: true
    toc-depth: 3
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(glue)

```

# **Practical One**

Fhatuwani Mokwenda

## **Part One**

```{r}
df <- airquality
```

Checking for nulls

```{r}
NaR = c()
for (rows in 1:nrow(df)){
  
  for (cols in 1: ncol(df)){
    
    if (is.na(df[rows, cols])){
      NaR <- c(NaR, rows)
      
    }
  }
  
}
print(NaR)
```

## **Part Two**

Finding the mean, sd, min and, max of the Temperature and Ozone

### **1. Looking at Temperature**

#### Mean

```{r}
mean(df$Temp)

```

#### Standard Deviation

```{r}
sd(df$Temp)

```

#### Min and Max

Note: Using Glue library for easy "fstrings"

```{r}
min = min(df$Temp)
max = max(df$Temp)

glue("max = {max}")
glue("min = {min}")
```

### **2. Looking at Ozone**

Have to remove NA rows first from df

```{r}
new.df <- df[-NaR,]

```

#### Mean

```{r}
mean(new.df$Ozone)
```

#### Standard Deviation

```{r}
sd(new.df$Ozone)
```

#### Min and Max

Note: Using Glue library for easy "fstrings"

```{r}
min = min(new.df$Ozone)
max = max(new.df$Ozone)

glue("max = {max}")
glue("min = {min}")
```

## **Part Three**

Linear regression, but instead of using lm()... we're using vector calculations!!!

```{r}
# predictor X -- no intercepts
x <- as.matrix(cars$speed)
# response Y
y <- as.matrix(cars$dist)

# Add a column of 1s to the predictor
X <- cbind(1, x)

# beta calculation
beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y
beta_hat

```

## Part Four

Checking **Part Three**

```{r}
coef(lm(dist~speed, data = cars))
```

We get the same thing !!!!!!!!

Thats it.
